Parameter estimation

### 2023-03-09 SFE-article/Sections/Parameter_estimation.tex
08.03
09.03

\documentclass[../Parameter_fitting.tex]{subfiles}
\graphicspath{{\subfix{../Figures/}}}
\begin{document}

Conceptually, the unobservable error $\epsilon(t)$ is added to the deterministic model output, $y(t)$, to give the observable dependent variable $Y(t)$, for example the experimental process output. For discrete observations:

{\footnotesize
	\begin{equation}
		Y(t_i) = y(t_i) + \epsilon(t_i)
\end{equation} }
and for continuos variables:

{\footnotesize
	\begin{equation}
		Y(t) = y(t) + \epsilon(t)
\end{equation} }

%If the estimated parameter $\hat{\theta}$ replaces the model parameter $\theta$, the residue error is $\textbf{E} = Y(t) - \hat{Y}(t)$, where $\hat{Y}(t)$ is the predicted value of $Y_i$ at $y_i$. 
The objective in parameter estimation is to obtain the "best" estimate of $\theta$ based on the continuous observations $Y(t)$ or the discrete observations $Y(t_i)$. 
\sout{
	We assume that $x(t)$ is given \textit{a priori} (for example as a solution of process model) but $\theta$ is to be estimated over the time interval $0 \leq t \leq t_n$ from discrete observation described by the following relation:}

{\footnotesize
	\begin{equation}
		Y(t_i) = \textbf{h}(t_i)y(t_i) + \boldsymbol{\epsilon}(t_i) \qquad 1 \leq i \leq n
\end{equation} }
\sout{where $ Y(t_i)$ is an $n \times 1$ column vector, $\textbf{h}(t_i)$ is an $n \times v$ matrix given \textit{a priori}, and $ \boldsymbol{\epsilon} (t_i)$ is an $n \times 1$ column vector ("noise" vector) whose elements are the unobservable errors. The above equation can be written in general form as:}

{\footnotesize
	\begin{equation} \label{EQ: Measurment_noise}
		Y(t_i) = y(\theta, t_i) + \epsilon(t_i)
\end{equation}}
where $y$ is the output model, which is a function of $\theta$ and time $t_i$.

Because of the difficulty of obtaining analytical solutions to the deterministic process model, experiments have been arranged whereby the vector of derivatives $d Y(t_i)/dt$ is measured rather than $Y(t_i)$ itself. In such cases, it is assumed that the unobservable error is added to the deterministic derivative $dy(\theta, t_i)/dt$ as follows:

{\footnotesize
	\begin{equation}
		\cfrac{d Y(t_i)}{dt} = \cfrac{dy(\theta, t_i)}{dt} + \epsilon(t_i)
\end{equation} }

\subsection{Least squares estimation}
A least squares parameter estimation does not require prior knowledge of the distribution of unobservable errors, yield  unbiased estimates, \sout{that is, $ \mathbb{E}\{\hat{Y}(t) = y(t) \} $,} and results in the minimum variance among all linear unbiased estimators. If the observations $ Y$ for the model responses are continuous functions of time from $t = 0$ to $t = t_f$, the Markov (or "rigorous least squares") criterion is to minimize:

@@

If $\Gamma$ is a diagonal matrix, $\phi$  becomes a "weighted least squares" criterion; if $\Gamma = \sigma^2 \textbf{I}$ where $\sigma$ is a standard deviation., $\phi$ is the "ordinary least squares" criterion.

\subsection{Weighted sum of squares - Cumulative data}

@@

The \sout{best} unbiased estimate of the Model B is obtained by minimizing the weighted sum of squares:

@@

Maximum likelihood estimation (MLE) is a method of estimating the parameters of an assumed probability distribution, given some observed data. This is achieved by maximizing a likelihood function so that, under the assumed statistical model, the observed data is most probable. The MLE has the desirable characteristics of asymptotic efficiency and normality. Each time it has been associated with the (joint) normal distribution because of mathematical convenience. Consider the joint probability density function (the likelihood function) $p(\theta |y)$ for random sample $y = y(t_1),y(t_2),...,y(t_n)$, parameter $\theta$. If a maximum of this function over all choices of $\theta$ can be found, the estimates so obtained are maximum likelihood estimates. The conditions at the maximum can be evolved incorporating prior information as follows.

@@

The numerator of the right-hand side od Equation \ref{EQ: Posterior_distribution} using Equation \ref{EQ: Probabiltiy_dependet_a} becomes

@@

By collecting a random sample of values of the random variable $y$ and selecting the values of $\theta$ that maximize the likelihood function $L\left( \theta | y \right)$, a function described in chapter \ref{CH: Bayes} in connection with Bayes' theorem. Such estimators, $\hat{\theta}$, are known as maximum likelihood estimators. In effect, the methods selects those values of $\theta$ that are at least as likely to generate the observed sample as any other set of values of the parameters if the probability density of the random variable $y$ were to be extensively simulated through use of the probability density $p\left( y|\theta \right)$. In making a maximum likelihood estimate, we assume that the form of the probability density (\sout{only the $\theta$ need be determined}) and that all possible values of $\theta$ are equally likely before experimentation.

@@

		{\footnotesize
			\begin{equation}
				\ln L = -n \ln \sqrt{2 \pi} - \cfrac{n}{2} \ln \sigma
				- \cfrac{ \sum_{i=1}^{n} \left[  Y(t_i) - y(\theta , t_i) \right]^2 }{2 \sigma^2}
			\end{equation}
		}

@@

@@

		\sout{By carrying out this operation, under fairly unrestrictive conditions,} 
		It can be shown that as $n$ approaches infinity the maximum likelihood estimates have the desirable asymptotic properties:

>>>>

\documentclass[../Article_Model_Parameters.tex]{subfiles}
\graphicspath{{\subfix{../Figures/}}}
\begin{document}

Conceptually, the unobservable error $\epsilon(t)$\todo[fancyline]{The same symbol as void fraction} is added to the deterministic model output, $y(t)$ (equation \ref{Model_measurment_1}), to give the observable dependent variable $Y(t)$ (for example results of an experiments). For discrete observations:

{\footnotesize
	\begin{equation*}
		Y(t_i) = y(t_i) + \epsilon(t_i)
\end{equation*} }
and for continuos variables:

{\footnotesize
	\begin{equation*}
		Y(t) = y(t) + \epsilon(t)
\end{equation*} }

%If the estimated parameter $\hat{\theta}$ replaces the model parameter $\theta$, the residue error is $\textbf{E} = Y(t) - \hat{Y}(t)$, where $\hat{Y}(t)$ is the predicted value of $Y_i$ at $y_i$. 
The objective in parameter estimation is to obtain the "best" estimate of $\theta$ based on the continuous observations $Y(t)$ or the discrete observations $Y(t_i)$. The above equation can be written in general form as:

{\footnotesize
	\begin{equation} \label{EQ: Measurment_noise}
		Y(t_i) = y(\theta, t_i) + \epsilon(t_i)
\end{equation}}

where $\theta$ is set of parameters to be estimated from the data. $\theta$ is a subset of parameter space $\Theta$, which contain all parameters of the model.

Because of the difficulty of obtaining analytical solutions to the deterministic process model, experiments have been arranged whereby the vector of derivatives $d Y(t_i)/dt$ is measured rather than $Y(t_i)$ itself. In such cases, it is assumed that the unobservable error is added to the deterministic derivative $dy(\theta, t_i)/dt$ (equation \ref{Model_measurment_2}) as follows:

{\footnotesize
	\begin{equation*}
		\cfrac{d Y(t_i)}{dt} = \cfrac{dy(\theta, t_i)}{dt} + \epsilon(t_i)
\end{equation*} }

If unobservable error in the first observation is designated as $\epsilon_1$, the second error observation $\epsilon_2'$ includes $\epsilon_1$ plus a random component introduced side from $\epsilon_1$, or $\epsilon_2' = \epsilon_1 + \epsilon_2$. The error in the third observation is $\epsilon_3' = \epsilon_1 + \epsilon_2 + \epsilon_3$, and so forth. \citet{Mandel1957} distinguished between the usually assumed type of independent measuring error in the dependent variable and a 'cumulative' or interval error in which each new observation includes the error of the previous observations. Cumulative errors, arising because of the fluctuations as a function of time in the process itself due to small changes in operating, are not independent - only the differences in measurement from one period to the next are independent.

\subsection{Least squares estimation}
A least squares parameter estimation does not require prior knowledge of the distribution of unobservable errors, yield  unbiased estimates, and results in the minimum variance among all linear unbiased estimators. If the observations $ Y$ for the model responses are continuous functions of time from $t = 0$ to $t = t_f$, the Markov (or "rigorous least squares") criterion is to minimize:

@@

If $\Gamma$ is a diagonal matrix, $\phi$  becomes a "weighted least squares" criterion; if $\Gamma = \sigma^2 \textbf{I}$ where $\sigma$ is a standard deviation, $\phi$ is the "ordinary least squares" criterion.

\iffalse
\subsection{Weighted sum of squares - Cumulative data}

@@

The unbiased estimate of the Model B is obtained by minimizing the weighted sum of squares:

@@

Maximum likelihood estimation (MLE) is a method of estimating the parameters of an assumed probability distribution, given some observed data. This is achieved by maximizing a likelihood function so that, under the assumed statistical model, the observed data is most probable. The MLE has the desirable characteristics of asymptotic efficiency and normality. Each time it has been associated with the (joint) normal distribution because of mathematical convenience. Consider the joint probability density function (the likelihood function) $p \left(\theta | y(t_1),y(t_2),...,y(t_n)\right)$ for parameter $\theta$. If a maximum of this function over all choices of $\theta$ can be found, the estimates so obtained are maximum likelihood estimates. The conditions at the maximum can be evolved incorporating prior information as follows.

@@

\todo[]{Notation is a bit unclear}The numerator of the right-hand side od Equation \ref{EQ: Posterior_distribution} using Equation \ref{EQ: Probabiltiy_dependet_a} becomes

@@

By collecting a values of $y$ and selecting the values of $\theta$ that maximize the likelihood function $L\left( \theta | y \right)$, a function described in chapter \ref{CH: Bayes} in connection with Bayes' theorem\todo{Is the reference it previous chapter needed?}. Such estimators, $\hat{\theta}$, are known as maximum likelihood estimators. In effect, the methods selects those values of $\theta$ that are at least as likely to generate the observed sample as any other set of values of the parameters if the probability density of $y$ were to be extensively simulated through use of the probability density $p\left( y|\theta \right)$. In making a maximum likelihood estimate, we assume that the form of the probability density (\sout{only the $\theta$ need be determined}) and that all possible values of $\theta$ are equally likely before experimentation. \todo{Needed?}

@@

		{\footnotesize
			\begin{equation}
				\ln L = -\cfrac{n}{2} \left( \ln \sqrt{2 \pi} + \ln \sigma^2 \right)
				- \cfrac{ \sum_{i=1}^{n} \left[  Y(t_i) - y(\theta , t_i) \right]^2 }{2 \sigma^2}
			\end{equation}
		}

@@

		\todo{$x$ has not been defined}

@@

		It can be shown that as $n$ approaches infinity the maximum likelihood estimates have the desirable asymptotic properties:

### 2023-03-09 SFE-article/Sections/Thermo.tex
09.03

@@

    Carbon dioxide above its critical point attains a supercritical state and exhibits certain peculiar properties such as higher density as a liquid but with a lower viscosity comparable to a gas. The supercritical carbon dioxide (S-CO2) has a range of engineering applications.

    The thermodynamic properties of S-CO2 such as density, local speed of sound and specific heat capacity vary significantly for slight change in temperature and pressure due to real gas effects. Equation of state (EOS) which account these real gas effects is used to calculate the thermodynamic properties. In order to predict this real gas effects, the Peng-Robinson equation of state (P-R EOS) is used to compute properties of $CO_2$. The P-R EOS belongs to the specific class of thermodynamic models for modelling the pressure of a gas as a function of temperature and density and can be written as a cubic function of the molar volume (of the density). The P-R EOS is presented by equation \ref{EQ: PR}

@@

    The Figure \ref{fig: SFE_Properties_CP} show behaviour of the heat capacity of a supercritical fluid at constant pressure ($C_p$). The details of the calculations can be found in the appendix. In contrary to the density which varies monotonically, the specific heat shows very high levels in a narrow region. In the subcritical region, the phase transition is associated with an effective spike in the heat capacity (i.e., the latent heat). Approaching the critical point, the latent heat falls to zero but this is accompanied by a gradual rise in heat capacity in the pure phases near phase transition. At the critical point, the latent heat is zero but the heat capacity shows a diverging singularity. Beyond the critical point, there is no divergence, but rather a smooth peak in the heat capacity; the highest point of this peak identifies the Widom line. (ref: 10.1038/nphys1683, 10.3311/PPch.12871)

    In order to calculate thermodynamic properties from a real gas, the departure function for that property with respect to the chosen equation of state has to be evaluated. The departure function for any thermodynamic property of a real gas is defined as the difference in the value of that property determined from the chosen real gas equation of state and the value of the same property for an ideal gas under the same conditions of temperature and pressure. (0.1007/s10494-017-9872-4)

@@

        On the other hand, transport properties (viscosity and conductivity) are also important quantities required in engineering design for production, fluid transportation, and processing. According to (10.1007/BF00500714), there is no satisfactory theory of transport properties of real dense gases and liquids. The main difficulties in the study of transport properties are twofold: one is the inherent difficulties involved in accurate measurements, and the other is the complexity involved in theoretical treatments. Therefore, the generally used correlations of transport coefficient are either empirical or based on some theoretical foundation. Enskog (REF) developed a popular theory for the transport properties of dense gas based on the distribution function. However, the Enskog theory was proposed for rigid spherical molecules. For real gases, some modification is needed. Following the Enskog theory, many correlations have been proposed in the form of the reduced density and reduced temperature. The correlations of (10.1063/1.556013) and (10.1063/1.4977429) implemented and compared as presented on Figure \ref{fig: SFE_Properties_mu}. Eventually the NIST correlation was selected to be used. The viscosity formulation proposed by NIST consists of four contributions: (i) for the limit of zero density, (ii) for the initial density dependence, (iii) for the residual viscosity, and (iv) for the singularity of the viscosity at the critical point. The NIST correlation covers temperatures from 100 to 2000 $K$ for gaseous $CO_2$ and from 220 to 700 $K$ with pressures along the melting line up to 8000 $MPa$ for compressed and supercritical liquid states.

@@

        \begin{figure*}[H]
			\centering
			\includegraphics[trim = 1.5cm 4.0cm 2.5cm 10.0cm,clip,width=\textwidth]{Figures/MU.pdf}	
			\caption{Viscosity obtained based on different correlations}
            \label{fig: SFE_Properties_mu}
		\end{figure*}

        Similarly, several correlations for thermal conductivity of $CO_2$ were compared on Figure
        \ref{fig: SFE_Properties_kt}. The presented figures are focused around critical point, where the singularity is present. Similarities between specific heat and the thermal conductivity can be observed. Eventually, the NIST correlation were selected. The correlations are applicable for the temperature range from the triple point to 1100 $K$ and pressures up to 200 $MPa$.

        \begin{figure*}[H]
			\centering
			\includegraphics[trim = 1.5cm 2.5cm 1.5cm 2.0cm,clip,width=0.95\textwidth]{Figures/KT.pdf}	
			\caption{Thermal conductivity obtained based on different correlations}
            \label{fig: SFE_Properties_kt}
		\end{figure*}

>>>>

    %Carbon dioxide above its critical point attains a supercritical state and exhibits certain peculiar properties such as higher density as a liquid but with a lower viscosity comparable to a gas. The supercritical carbon dioxide (S-CO2) has a range of engineering applications.

@@

    The thermodynamic properties of supercritical $CO_2$ such as density, local speed of sound and specific heat capacity vary significantly for slight change in temperature and pressure due to real gas effects. Equation of state (EOS) which account these real gas effects is used to calculate the thermodynamic properties. In order to predict this real gas effects, the Peng-Robinson equation of state (P-R EOS) is used to compute properties of $CO_2$. Detail information about Peng-Robinson equation of state can be found in the work of \citet{Peng1976}, \citet{Elliott2011} or \citet{Pratt2001}. The P-R EOS belongs to the specific class of thermodynamic models for modelling the pressure of a gas as a function of temperature and density and can be written as a cubic function of the molar volume (of the density). The P-R EOS is presented by equation \ref{EQ: PR}

@@

    The Figure \ref{fig: SFE_Properties_CP} show behaviour of the heat capacity of a supercritical fluid at constant pressure ($C_p$). The details of the calculations can be found in the appendix. In contrary to the density which varies monotonically, the specific heat shows very high levels in a narrow region. In the subcritical region, the phase transition is associated with an effective spike in the heat capacity (i.e., the latent heat). Approaching the critical point, the latent heat falls to zero but this is accompanied by a gradual rise in heat capacity in the pure phases near phase transition. At the critical point, the latent heat is zero but the heat capacity shows a diverging singularity. Beyond the critical point, there is no divergence, but rather a smooth peak in the heat capacity; the highest point of this peak identifies the Widom line (as discussed by \citet{Simeoni2010} and \citet{Banuti2019}).

    In order to calculate thermodynamic properties from a real gas, the departure function for that property with respect to the chosen equation of state has to be evaluated. As presented by \citet{Elliott2011}, the departure function for any thermodynamic property of a real gas is defined as the difference in the value of that property determined from the chosen real gas equation of state and the value of the same property for an ideal gas under the same conditions of temperature and pressure. %(10.1007/s10494-017-9872-4)

@@

        On the other hand, transport properties (viscosity and conductivity) are also important quantities required in engineering design for production, fluid transportation, and processing. According to \citet{Sheng1989}, there is no satisfactory theory of transport properties of real dense gases and liquids. The main difficulties in the study of transport properties are twofold: one is the inherent difficulties involved in accurate measurements, and the other is the complexity involved in theoretical treatments. Therefore, the generally used correlations of transport coefficient are either empirical or based on some theoretical foundation. Enskog (\todo[fancyline]{Add ref}) developed a popular theory for the transport properties of dense gas based on the distribution function. However, the Enskog theory was proposed for rigid spherical molecules. For real gases, some modification is needed. Following the Enskog theory, many correlations have been proposed in the form of the reduced density and reduced temperature. The correlations of \citet{Fenghour1998} and \citet{Laesecke2017} implemented and compared as presented on Figure \ref{fig: SFE_Properties_mu}. The viscosity formulation proposed by NIST consists of four contributions: (i) for the limit of zero density, (ii) for the initial density dependence, (iii) for the residual viscosity, and (iv) for the singularity of the viscosity at the critical point. The NIST correlation covers temperatures from 100 to 2000 $K$ for gaseous $CO_2$ and from 220 to 700 $K$ with pressures along the melting line up to 8000 $MPa$ for compressed and supercritical liquid states.

@@

        \begin{figure*}[H]
			\centering
			\includegraphics[trim = 1.5cm 4.0cm 2.5cm 10.0cm,clip,width=\textwidth]{Figures/MU.pdf}	
			\caption{Viscosity obtained based on different correlations ({\color{red}Decrease space between figures})} 
            \label{fig: SFE_Properties_mu}
		\end{figure*}

        Similarly, several correlations for thermal conductivity of $CO_2$ were compared on Figure \ref{fig: SFE_Properties_kt}. The presented figures are focused around critical point, where the singularity is present. Similarities between specific heat and the thermal conductivity can be observed. The NIST correlation (\citet{Huber2016}) captures the singular behaviour of thermal conductivity around the critical point. The correlations are applicable for the temperature range from the triple point to 1100 $K$ and pressures up to 200 $MPa$.

        \begin{figure*}[H]
			\centering
			\includegraphics[trim = 1.5cm 2.5cm 1.5cm 2.0cm,clip,width=0.95\textwidth]{Figures/KT.pdf}	
			\caption{Thermal conductivity obtained based on different correlations }
            \label{fig: SFE_Properties_kt}
		\end{figure*}
